# Sequences and Series

## Sequences

- A **Sequence** is a list of numbers written in a definite order:
 \[a_{1}, a_{2}, a_{3}, a_{4}, ..., a_{n}\]
- An element of a sequence is called a term . The number $a_{1}$ is called the *first term*, $a_{2}$ is  the *second term*, and in general $a_{n}$ is called the $n^{th}$ *term*.
- As can be seen in the above sequence, for every positive integer $n$ there is a corresponding number $a_{n}$.Therefore a sequence can be defined as a function whose domain is the set of positive integers.
- But we usually write $a_{n}$  instead of the function notation $f(n)$ for the value of the function at the number $n$.
- When dealing with infinite sequences, each term $a_{n}$ will have a successor $a_{n+1}$
- **Notation:** The sequence $\{a_{1}, a_{2}, a_{3},...\}$ is also denoted by 
\[\{a_{n}\}\]  or \[\{a_{n}\}_{n=1}^{\infty}\]
- Some sequences can be defined by giving a formula for the $n$th term.
- In the following examples we give three descriptions of the sequences:
    i. by using the preceding notation
    ii. by using the defining formula
    iii. by writing out the terms of the sequence
- Notice that $n$ doesn't have to start at 1.

a) $\{\frac{n}{n+1}\}_{n=1}^{\infty} \hspace{5em} a_{n} = \frac{n}{n+1}\hspace{5em}\{\frac{1}{2},\frac{2}{3}, \frac{3}{4}, \frac{4}{5},..., \frac{n}{n+1},... \}$

b) $\{\frac{(-1)^n(n+1)}{3^{n}}\}_{n=1}^{\infty} \hspace{5em} a_{n} = \frac{(-1)^n(n+1)}{3^{n}}\hspace{5em}\{-\frac{2}{3},\frac{3}{9}, -\frac{4}{27}, \frac{5}{81},..., \frac{(-1)^n(n+1)}{3^{n}},... \}$

c) $\{\sqrt{n-3}\}_{n=3}^{\infty} \hspace{5em} a_{n} = \sqrt{n-3}, n\geq3\hspace{5em}\{0,1,\sqrt{2}, \sqrt{3},..., \sqrt{n-3},...\}$

d) $\{cos\frac{n\pi}{6}\}_{n=0}^{\infty} \hspace{5em} a_{n} = cos\frac{n\pi}{6}, n\geq 0 \hspace{5em}\{1, \frac{\sqrt{3}}{2},\frac{1}{2}, 0,...,cos\frac{n\pi}{6},...\}$


<!--**Tutorial 1-->

## Series

- A series is the sum of a number of terms of a sequence.
- When writing series, the shorthand $\sum$ notation is used to represent the sum of a number of terms having a common form.
- The series $f(1)+ f(2)+\dots+ f(n−1)+ f(n)$ can be written as
\[\sum_{r=1}^nf(r).\]

### Arithmetic sequences and series

- A sequence in which each term after the first term is obtained from the preceding term by adding a fixed number (Common difference), is called an arithmetic sequence or Arithmetic Progression.

- The sequence defined by
\[u_1=a \text{ and } u_n=u_{n-1} + d \text{ for } n\geq2\]
gives
\[a,a+d, a+2d,\dots\]
- The $n$th term (i.e. the solution) is given by $u_n = a + (n −1)d$.
- This is the **arithmetic sequence** with first term $a$ and common difference $d$


- The **arithmetic series** with $n$ terms 
\[a+(a+d)+(a+2d)+\dots+ [a+(n-1)d]\]
has sum
\[S_n = \frac{n}{2}(\text{first term} + \text{ last term)}\]
\[S_n = \frac{n}{2}(2a+(n-1)d)\]


### Geometric sequences and series


- The sequence defined by
\[u_1=a \text{ and } u_n=ru_{n-1} \text{ for } n\geq2\]
gives
\[a,ar, ar^2,\dots\]
- The $n$th term is given by $u_n = ar^{n −1}$.
- This is the **geometric sequence** with first term $a$ and common ratio $r$.

- The **geometric series** with $n$ terms
\[a+ar+ar^2+\dots+ ar^{n-1}\]
has sum
\[S_n = \frac{a(1-r^n)}{1-r} \text{ or }  \frac{a(r^n-1)}{r-1} \text{ for } r\neq1\]




**Important results relating to the** $\sum$ **notation**

(1) 
$$\sum_{r=1}^n\{f(r)+ g(r)\} = \{f(1)+ g(1)\}+ \{f(2)+ g(2)\}+\dots + \{f(n)+ g(n)\}$$
$$= \{f(1)+ f(2)+ \dots+ f(n)\}+ \{g(1)+ g(2)+\dots +g(n)\}$$
$$= \sum_{r=1}^nf(r)+ \sum_{r=1}^ng(r)$$

(2)
$$\sum_{r=1}^naf(r)= af(1)+ af(2)+ \dots+ af(n)$$ where $a$ is a constant.
$$=a\{f(1)+ f(2)+ \dots+ f(n)\}$$
$$=a\sum_{r=1}^nf(r)$$

<!-- $n\sum_{r=1}^nf(r) = \sum_{r=1}^nf(r)$ also, since n is a fixed quantity,
Note that
not a variable, in the summation-->

(3)

(a) $\sum_{r=1}^nr=\frac{n(n+1)}{2}$

(b) $\sum_{r=1}^nr^2=\frac{n}{6}(n+1)(2n+1)$

(c) $\sum_{r=1}^nr^3=\frac{n^2}{4}(n+1)^2$ or $\left[\frac{n(n+1)}{2}\right]^2$

(d) $\sum_{r=1}^n1=(1+1+\dots+1) = n$

<!-- Since a '1' is wanted for each of r =1,2,...,n
This result is easily overlooked, and often incorrectly written as
$\sum_{r=1}^n1=1$-->


<!-- Tutorial 2-->

###  Methods of proof

  1. Mathematical induction
  2. The difference method

#### Mathematical induction

<!-- Acharjya, D. P. (2009). Fundamental approach to discrete mathematics. New Age International. page 8-->


- Let $n$ be a natural number. Then the aim is to show that some statement $P(n)$ involving $n$ is true for any $n$. 
- The following steps are used in Mathematical induction
   
   1. Let $P(n)$ be a statement
   2. Show that the statement is true for  $P(1)$ and $P(2)$. (*i.e.* $P(n)$ is true for $n=1$ and $n=2.$)
   3. Assume that $P(k)$ is true (*i.e.* $P(n)$ is true for $n=k$).
   4. Show that $P(k+1)$ follows from $P(k).$
   
#### The difference method

- The process of proof by induction is a powerful mathematical tool. However it has the disadvantage that, in order to employ the method it requires the answer. 
- There are, however, direct methods of proof such as the method of differences, or the difference method.
- The difference method can be summarised as follows,

\[\sum_{r=1}^n \left\{f(r)-f(r-1) \right\} = f(n) - f(0)\]

where $f$ is any function suitably defined on the non-negative integers.

- This is also know  as the fundamental theorem of summation:


##  Infinite Sequences

- When dealing with infinite sequences, each term $a_{n}$ will have a successor $a_{n+1}$

**Definition: Convergence and Divergence**

A sequence $\{a_n\}$ has the **limit** $L$ and we write 

\[\lim_{n\rightarrow \infty}a_n=L \text{ or } a_n \rightarrow L \text{ as } n\rightarrow \infty\]

if we can make the term $a_n$ as close to L as $n$ becomes sufficiently large.

If $\lim_{n\rightarrow \infty}a_n$ exists, we say the sequence **converges** (or is **convergent**).

If $\lim_{n\rightarrow \infty}a_n$ does not exist, we say the sequence **diverges** (or is **divergent**).


```{r boxC21, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```


### Limit Laws for Sequences

If $\{a_n\}$ and $\{b_n\}$ are convergent sequences and $c$ is a constant, then

a) $\lim_{n \to \infty}(a_n+ b_n)= \lim_{n \to \infty}a_n+ \lim_{n \to \infty}b_n$

b) $\lim_{n \to \infty}(a_n- b_n)= \lim_{n \to \infty}a_n- \lim_{n \to \infty}b_n$

c) $\lim_{n \to \infty}ca_n= c \;\lim_{n \to \infty}a_n$

d) $\lim_{n \to \infty}c= c$

e) $\lim_{n \to \infty}(a_nb_n)= \lim_{n \to \infty}a_n. \lim_{n \to \infty}b_n$

f) $\lim_{n \to \infty}\frac{a_n}{b_n}= \frac{\lim_{n \to \infty}a_n} {\lim_{n \to \infty}b_n} \text{ if } \lim_{n \to \infty}{b_n} \neq 0$

g) $\lim_{n \to \infty}(a_n)^p= [\lim_{n \to \infty}a_n]^p \text{ if } p>0 \text{ and } a_n>0$


**Squeeze Theorem for Sequences**

If $a_n \leq b_n \leq c_n$ for $n\geq n_0$ and $\lim_{n \to \infty}a_n= \lim_{n \to \infty}c_n= L,$ then $\lim_{n \to \infty}b_n=L$

**Theorem:** 

$$\text{If }\;\; \lim_{n \to \infty}|a_n|=0, \text{ then } \;\;\lim_{n \to \infty}a_n=0.$$


**Theorem**

The sequence $\{r^n\}$ is convergent if $-1 <r\leq1$ and divergent for all other values of $r$.

\begin{equation}
\lim_{n \to \infty}r^n=
\begin{cases} 
0 & \text{if } -1<r<1\\
1 & \text{if } r=1
\end{cases}
\end{equation}

**Definition**

A sequence $\{a_n\}$ is called **increasing** if $a_n<a_{n+1}$ for all $n \geq 1,$ that is, $a_1<a_2<a_3<\dots.$ It is called **decreasing** if $a_n>a_{n+1}$ for all $n\geq 1.$ It is called **monotonic** if it is either increasing or decreasing.

<!-- CalcII_Seq_Series.pdf page 17 stewart page 708

Discuss tutorial question 6 and 7-->

\newpage

**Definition**

- If there exists a number $m$ such that $m \leq a_n$ for every $n$ we say the sequence is **bounded below**. The number $m$ is sometimes called a *lower bound* for the sequence.

- If there exists a number $M$ such that $a_n \leq M$ for every $n$ we say the sequence is **bounded above**. The number $M$ is sometimes called an *upper bound* for the sequence.

- If the sequence is both bounded below and bounded above we call the sequence **bounded**.

<!-- Discuss tutorial question 8-->

##  Infinite  Series


- Consider the infinite sequence $\{a_n\}_{n=1}^\infty$.
- Consider the *partial sums*

\begin{align*}
   s_1=a_1 \\
   s_2=a_1+a_2 \\
   s_3=a_1+a_2+a_3 \\
   s_4=a_1+a_2+a_3+a_4 \\
   &\vdots\\
   s_n=a_1+a_2+a_3+a_4+\dots +a_n =\sum_{i=1}^na_i
\end{align*}

- These partial sums form a new sequence $\{s_n\}_{n=1}^\infty,$ which may or may not have a limit. 

- Consider the limit of the sequence of partial sums, $\{s_n\}_{n=1}^\infty.$

$$\lim_{n\to\infty}s_n= \lim_{n\to\infty}\sum_{i=1}^na_i=\sum_{i=1}^\infty a_i$$

- This $\sum_{i=1}^\infty a_i$ ($=a_1+a_2+a_3+a_4+\dots +a_n \dots$) is called an **infinite series**

\newpage 
**Definition**

Consider the series $\sum_{n=1}^\infty a_n= =a_1+a_2+a_3+\dots.$ Let $s_n$ denote its $n$th partial sum:

$$s_n= \sum_{i=1}^n a_i= a_1+a_2+\dots +a_n$$

If the sequence $\{s_n\}$ is convergent and $\lim_{n\to \infty}s_n=s$ exist as a real number, then the series $\sum_{i=1}^\infty a_i$ is called **convergent** and we write,

$a_1+a_2+\dots +a_n+\dots=s$ or $\sum_{n=1}^\infty a_n=s$

The number $s$ is called the **sum of the series**. 
If the sequence of partial sums is divergent then the infinite series is also
called **divergent**.

<!-- stewart page 714 note before example 1-->

**Geometric Series**

<!-- Stewart page 714 example 1-->

- An important example of an infinite series is the *geometric series*

$$\sum_{n=1}^\infty ar^{n-1}=a+ar+ar^2+\dots$$
is convergent if $|r|<1$ and its sum is 
$$\sum_{n=1}^\infty ar^{n-1}= \frac{a}{1-r}\qquad\qquad|r|<1$$

If $|r|\geq1,$ the geometric series is divergent.

<!-- tutorial question 9 and 10-->

\newpage 

<!-- stewart page 718-->
**Theorem**

If the series $\sum_{n=1}^\infty a_n$ is convergent, then $\lim_{n\to \infty}a_n=0.$

*Proof*

```{r seqbox3, out.width='100%', fig.asp=.95, fig.align='center', fig.pos='h'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```
<!-- stewart page 718-->
**Note 1**

- Consider any series $\sum a_n.$
- This associates two sequences

  - the  sequence $\{s_n\}$ of its *partial sums* 
  - the sequence  $\{a_n\}$ of its *terms*.
- If $\sum a_n$ is convergent, then 

  - the limit of the sequence $\{s_n\}$  is $s$ (the sum of the series)
  - the limit of the sequence  $\{a_n\}$ is 0.

<!-- stewart page 718--> 
**Note 2**
- The converse of the above theorem is not true in general.

```{r seqbox4, out.width='100%', fig.asp=.5, fig.align='center', fig.pos='h'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

**The Test for Divergence**

If $\lim_{n\to \infty}a_n$ does not exist or if $\lim_{n\to \infty}a_n \neq0$, then the series $\sum_{n=1}^\infty a_n$ is divergent.

<!-- Tutorial Question11-->


**Theorem**


If $\sum a_n$ and $\sum b_n$ are convergent series, then 

a) $\sum_{n=1}^\infty ca_n= c\sum_{n=1}^\infty a_n \;\;$ $c$ is a constant.

b) $\sum_{n=1}^\infty (a_n+b_n)= \sum_{n=1}^\infty a_n+ \sum_{n=1}^\infty b_n$

c) $\sum_{n=1}^\infty (a_n-b_n)= \sum_{n=1}^\infty a_n- \sum_{n=1}^\infty b_n$

<!-- proof  stewart page 719-->

<!-- tutorial 12-->
Reading:

- Stewart, J., Clegg, D. K., & Watson, S. (2020). 'Infinite Sequences and Series', *Calculus: early transcendentals*. Cengage Learning.

- Acharjya, D. P. (2009). Fundamental approach to discrete mathematics. New Age International.

\newpage
\pagenumbering{arabic}
## Tutorial {-}

<!-- Stewart page 702-->
1. Find a formula for the general term $a_{n}$ of the sequence, assuming that the pattern of the first few terms continues.

(a) $\left\{\frac{3}{5},-\frac{4}{25}, \frac{5}{125}, -\frac{6}{625}, \frac{7}{3125},...\right\}$
(b) $\{1,1,2,3,5,8,13,21,\dots\}$
(c)  $\left\{\frac{1}{2}, \frac{1}{4}, \frac{1}{8}, \frac{1}{16}, \dots \right\}$
(d)  $\left\{2,7,12,17, \dots \right\}$
(e)  $\left\{1,-\frac{2}{3}, \frac{4}{9} , -\frac{8}{27}, \dots \right\}$

2. Show that 
\[\sum_{r=1}^n(6r^2+4r-1) = n(n+2)(2n+1)\]

<!--acharjya page 8-->

3. Show by method of induction

(a) $\sum_{r=1}^nr=\frac{n(n+1)}{2}$
(b) $\sum_{r=1}^nr^2=\frac{n}{6}(n+1)(2n+1)$
(c) $\sum_{r=1}^nr^3=\left[\frac{n(n+1)}{2}\right]^2$ <!--acharjya page 14-->
(d) $\sum_{r=1}^n\frac{1}{r(r+1)}=\frac{n}{n+1}$ <!--acharjya page 14-->

4. 

(a) By considering $n^3 − (n − 1)^3$ and similar expressions, find the formula for $\sum_{r=1}^nr^2$ in terms of $n$, assuming the results for
$\sum_{r=1}^nr$

(b) By considering $n^5 − (n − 1)^5$ and similar expressions, find the formula for $\sum_{r=1}^nr^4$ in terms of $n$, assuming the results for
$\sum_{r=1}^nr$, $\sum_{r=1}^nr^2$ and $\sum_{r=1}^nr^3.$

<!-- stewart page 706 Example 6)-->
5. Determine whether the sequence $a_n= (-1)^n$ is convergent or divergent.

<!-- stewart page 706 Example 7)-->
6. Determine whether the sequence converges or diverges. If it converges, find the limit

a) $a_n= n(n-1)$
b) $a_n= \frac{n}{n+1}$
c) $a_n = \frac{(-1)^n}{n}$ <!--CalcII_Seq_Series.pdf page 15-->
d) $a_n = \frac{4n^2+2}{n^2+n}$
e) $a_n= \frac{2^n}{3^{n+1}}$
f) $a_n= \frac{(-1)^{n-1}n}{n^2+1}$

<!-- stewart page 706 Example 8): Squeez theorem-->
7. Discuss the convergence of the sequence $a_n=n!/n^n,$ where $n!= 1\times2\times\dots\times n.$

<!-- CalcII_Seq_Series.pdf page 18(191)-->
8. Determine if the following sequences are monotonic  and/ or bounded

a) $\{-n^2\}_{n=0}^\infty$
b) $\{(-1)^{n+1}\}_{n=1}^\infty$
c) $\left\{ \frac{2}{n^2}\right\}_{n=5}^\infty$

9. Determine if the following series converges or diverges. If it converges determine its value.

a) $\sum_{n=1}^\infty n$  <!--CalcII_Seq_Series.pdf page 30 (203) exampl 1-->
b) $\sum_{n=1}^\infty (-1)^n$ <!--CalcII_Seq_Series.pdf page 31 (204) example 3-->
c) $\sum_{n=1}^\infty \frac{1}{3^{n-1}}$ <!--CalcII_Seq_Series.pdf page 32 (203) exampl 1-->
d) $5-\frac{10}{3}+\frac{20}{9}- \frac{40}{27}+\dots$ <!-- stewart page 715 example 2-->
e) $\sum_{n=1}^\infty 2^{2n}3^{1-n}$ <!-- stewart page 716 example 3-->
f) $\sum_{n=1}^\infty x^n,$ where $|x| <1$ <!-- stewart page 716 example 5-->
g) $\sum_{n=1}^\infty\frac{1}{n(n+1)}$ <!-- stewart page 717 example 6-->

<!-- stewart page 717 example 7-->
10. Show that harmonic series 
$$\sum_{n=1}^\infty \frac{1}{n} = 1+\frac{1}{2}+ \frac{1}{3} + \frac{1}{4}+\dots$$ is divergent.

<!-- stewart page 717 example 8-->
11.  Show that the series $\sum_{n=1}^\infty\frac{n^2}{5n^2+4}$ diverges.



12. Determine whether the series is convergent or divergent. If it is convergent, find its sum. 

a) $\sum_{n=1}^\infty\left(\frac{3}{n(n+1)+ }+\frac{1}{2^n}\right)$ <!-- stewart page 719 example 9-->
b) $\sum_{n=1}^\infty\left(\frac{3}{n(n+3)+ }+\frac{5}{4^n}\right)$ <!-- stewart page 720 Question 33 example 9 answer 7/2-->
c)  $\sum_{n=1}^\infty \frac{1}{n^2+3n+2}$ <!--CalcII_Seq_Series.pdf page 42 (215) example 3 ans 1-->
d) $\sum_{n=1}^\infty \frac{1}{n^2+4n+3}$ <!--CalcII_Seq_Series.pdf page 42 (215) example 4 answer 5/12-->
e) $\sum_{n=1}^\infty 9^{-n+2}4^{n+1}$ <!--CalcII_Seq_Series.pdf page 39 (212) example 4 answer 1296/5-->
f) $\sum_{n=1}^\infty \frac{(-4)^{3n}}{5^{n-1}}$ <!--CalcII_Seq_Series.pdf page 39 (212) example 4 answer 1296/5-->
g) $\sum_{n=1}^\infty \left(\frac{1}{n^2+4n+3}- 9^{-n+2}4^{n+1}\right)$ <!--CalcII_Seq_Series.pdf page 44 (217) example 5 answer 3863/15-->