# Introduction to Logic

\pagenumbering{arabic}

## Statement (Proposition)

- A statement is a declarative sentence which is either true or false **but not both**
- A proposition is a collection of declarative statements that has either a truth value "true” or a truth value "false".
- The truth value True and False are denoted by the symbols **T** and **F**, respectively.
- Sometimes it is also denoted by **1** (stands for true) and **0** (stands for false).
- Since it depends on only two possible truth values, it is known as *two-values logic* of *bi-values logic*.

*Example 1*

Consider the following sentences and select the statements

  (a) Man is Mortal <!-- (subject to death),it returns truth value “TRUE”-->
  (b) 12 plus 9 is equal to 1 <!-- it returns truth value “FALSE-->
  (c) Moon rises in the east <!-- The Sun, the Moon, the planets, and the stars all rise in the east and set in the west.-->
  (d) 3 is less than four
  (e) Please sit down
  (f) $y$ is a cat <!--  It is because unless we give a specific value of A, we cannot say whether the statement is true or false.-->
  (g) 14 is a composite number <!--A composite number is a positive integer that can be formed by multiplying two smaller positive integers. Equivalently, it is a positive integer that has at least one divisor other than 1 and itself. ... For example, the integer 14 is a composite number because it is the product of the two smaller integers 2 × 7.
  
  A composite number has more than two factors. The number 1 is neither prime nor composite. The prime numbers between 2 and 31 are 2, 3, 5, 7, 11, 13, 17, 19, 23, 29 and 31 since each of these numbers has only two factors, itself and 1-->
  (f) *Chaddy* is a nice dog.

  
## Propositional Variables

-  Every statement in propositional logic consists
of **propositional variables** combined via
**propositional connectives**.
- Propositional variables are usually denoted by capital English letters, such as P, Q, R etc.
- Each variable can take one of two values: **T** or **F**.

## Logical Connectives

<!--https://www.tutorialspoint.com/discrete_mathematics/discrete_mathematics_propositional_logic.htm#:~:text=Propositional%20Logic%20is%20concerned%20with,or%20in%20a%20composite%20manner.

http://web.stanford.edu/class/archive/cs/cs103/cs103.1152/lectures/07/Slides07.pdf
-->

- We use **logical connectives** to connect  the propositional variables (several statements into a single statement). 
- The most basic and fundamental connectives are Negation, Conjunction and Disjunction.

### Negation

- Negation is the simplest common example of a truth-functional operation.
- The negation of a statement is also a statement.
- We use the connective **Not** for negation. 
- If $P$ be a statement, then the negation of $P$ is denoted by $\neg P$.
- $P$ and $\neg P$ has opposite truth values.
- The relationship between the truth values of $P$ and $\neg P$  can be made explicit by a diagram called a **truth table**

**Truth Table (Negation)**

$P$   | $\neg P$
------|------------
$T$   |   $F$
$F$   |   $T$


### Conjunction

- The conjunction of two statements, $P$ and $Q$ is also a statement.
- We use the connective **And** ($\land$) for conjunction.
- It is denoted by $P \land Q$

**Truth Table (Conjunction)**

$P$   | $Q$  | $(P \land Q)$
------|----- |------------
$T$   |  $T$ |  $T$
$T$   |  $F$ |  $F$
$F$   |  $T$ |  $F$
$F$   |  $F$ |  $F$

**Rule:**  $(P \land Q)$ is true if both $P$ and $Q$ are true, otherwise false.

- In order to make  $(P \land Q)$ true, $P$ and $Q$ have to be **simultaneously** true


### Disjunction

- The disjunction of two statements, $P$ and $Q$ is also a statement.
- We use the connective **Or** ($\lor$) for disjunction
- It is denoted by $P \lor Q$

**Truth Table (Disjunction)**

$P$   | $Q$  | $(P \lor Q)$
------|----- |------------
$T$   |  $T$ |  $T$
$T$   |  $F$ |  $T$
$F$   |  $T$ |  $T$
$F$   |  $F$ |  $F$

**Rule:**  $(P \lor Q)$ is true if either $P$ or $Q$ is true and it is false when both $P$ and $Q$ are false.

- $(P \lor Q)$ shall stand for "$P$ or $Q$ or both".

## Conditional

- In mathematics, expressions of the form "IF $P$ then $Q$" occur so often.
- This can be expressed in any one of the following forms.
    a) If $P$, then $Q$
    b) $P$ only if $Q$
    c) $P$ implies $Q$
    d) $Q$ if $P$
- Therefore, it is important to understand the corresponding truth-functional operation.
- Let $P$ and $Q$ be any two statements.
- Then the statement $P \rightarrow Q$ is called a conditional statement.
- In an implication $P \rightarrow Q$, $P$ is called the *antecedent* (hypothesis) and $Q$ is called the *consequent* (conclusion).

**Truth Table (Conditional)**

$P$   | $Q$  | $(P \rightarrow Q)$
------|----- |------------
$T$   |  $T$ |  $T$
$T$   |  $F$ |  $F$
$F$   |  $T$ |  $T$
$F$   |  $F$ |  $T$

**Rule:**  An implication (conditional) $(P \rightarrow Q)$ is False only when the hypothesis ($P$) is true and conclusion ($Q$) is false, otherwise True.

## Bi-Conditional

- Let $P$ and $Q$ be any two statements.
- Then the statement $P \leftrightarrow Q$ is called a bi-conditional statement.
- This can be expressed in any one of the following forms.
    a) $P$ if and only if $Q$
    b) $P$ is necessary and sufficient of $Q$
    c) $P$ is necessary and sufficient for $Q$
    d) $P$ implies $Q$ and $P$ is implied by $Q$
- The bi-conditional (double implication) $P \leftrightarrow Q$ can be defined as 

$$P \leftrightarrow Q: (P \rightarrow Q) \land (Q \rightarrow P)$$

\newpage

**Truth Table (Bi-Conditional)**

$P$   | $Q$  |  $(P \rightarrow Q)$ | $(Q \rightarrow P)$ |  $(P \leftrightarrow Q)$
------|----- |----------------------|---------------------|-----------
$T$   |  $T$ |  $T$   | $T$    | $T$
$T$   |  $F$ |  $F$   | $T$    | $F$
$F$   |  $T$ |  $T$   | $F$    | $F$
$F$   |  $F$ |  $T$   | $T$    | $T$  
    
**Rule:**  $(P \leftrightarrow Q)$ is True only when both $P$ and $Q$ have identical truth values, otherwise false. 

<!-- From the truth table us is clear that $(P \leftrightarrow Q)$ has the truth value T whenever both P and Q have identical truth values-->

## Converse

- Let $P$ and $Q$ be any two statements.
- The converse statement of the conditional  $P \rightarrow Q$ is given as  $Q \rightarrow P$

```{r boxC31, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

## Inverse

- Let $P$ and $Q$ be any two statements.
- The inverse statement of the conditional  $(P \rightarrow Q)$ is given as  $(\lnot P \rightarrow \lnot Q)$

```{r boxC32, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

## Contra Positive

- Let $P$ and $Q$ be any two statements.
- The contra positive statement of the conditional  $(P \rightarrow Q)$ is given as  $(\lnot Q \rightarrow \lnot P)$

```{r boxC33, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```


**Truth Table (Contra Positive)**

$P$   | $Q$  |  $P \rightarrow Q$ | $\lnot Q$ | $\lnot P$ | $(\lnot Q \rightarrow \lnot P)$ 
------|----- |------------|-----------|-----------|--------
$T$   |  $T$ |  $T$   | $F$    | $F$ | $T$
$T$   |  $F$ |  $F$   | $T$    | $F$ | $F$
$F$   |  $T$ |  $T$   | $F$    | $T$ | $T$
$F$   |  $F$ |  $T$   | $T$    | $T$ | $T$ 
    
- From the truth table it can be seen that both conditional $P \rightarrow Q$  and contra positive  $(\lnot Q \rightarrow \lnot P)$ have same truth values.

## Exclusive OR

- Let $P$ and $Q$ be any two statements.
- The exclusive OR of two statements $P$ and $Q$ is denoted by $(P \overline{\lor} Q)$.
- We use the connective XOR for exclusive OR
- The exclusive OR $(P \overline{\lor} Q)$ is true if either $P$ or $Q$ is True but not both.
- The exclusive OR is also known as exclusive disjunction.

the conditional  $(P \rightarrow Q)$ is given as  $(\lnot Q \rightarrow \lnot P)$

**Truth Table (Exclusive OR)**

$P$   | $Q$  | $(P \overline{\lor}Q)$
------|----- |------------
$T$   |  $T$ |  $F$
$T$   |  $F$ |  $T$
$F$   |  $T$ |  $T$
$F$   |  $F$ |  $F$
 
**Rule:** $(P \overline{\lor}Q)$ is true if either $P$ or $Q$ is True but not both, otherwise false.


## NAND

- The word NAND stands for NOT and AND.
- The connective NAND is denoted by the symbol $\uparrow$.
- Let $P$ and $Q$ be any two statements.
- The NAND of $P$ and $Q$ is denoted by $(P \uparrow Q)$.
- The NAND, $(P \uparrow Q)$ can be defined as 

$$(P\uparrow Q)\equiv\lnot(P \land Q)$$

**Truth Table (NAND)**

$P$   | $Q$  | $(P \uparrow Q)$
------|----- |------------
$T$   |  $T$ |  $F$
$T$   |  $F$ |  $T$
$F$   |  $T$ |  $T$
$F$   |  $F$ |  $T$
 
**Rule:** $(P \uparrow Q)$ is True if *either* $P$ or $Q$ is false, otherwise False


## NOR

- The word NOR stands for NOT and OR.
- The connective NOR is denoted by the symbol $\downarrow$.
- Let $P$ and $Q$ be any two statements.
- The NOR of $P$ and $Q$ is denoted by $(P \downarrow Q)$.
- The NOR, $(P \downarrow Q)$ can be defined as 

$$(P\downarrow Q)\equiv\lnot(P \lor Q)$$

**Truth Table (NOR)**

$P$   | $Q$  | $(P \downarrow Q)$
------|----- |------------
$T$   |  $T$ |  $F$
$T$   |  $F$ |  $F$
$F$   |  $T$ |  $F$
$F$   |  $F$ |  $T$
 
**Rule:** $(P \downarrow Q)$ is True only when both $P$ or $Q$ are false, otherwise False

## Tautology

-  If the truth values of a composite statement are always true, irrespective of the truth values of the atomic (individual) statements, then it is a **tautology**.

*Example*

- The composite statement $(P\land (P \rightarrow Q)) \rightarrow Q$ is a tautology.
- Verify this using a truth table with composite statement as  $(P\land (P \rightarrow Q)) \rightarrow Q$

```{r boxC35, fig.cap='', out.width='100%', fig.asp=.7, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

## Contradiction

- If the truth values of a composite statement are always false, irrespective of the truth values of the atomic (individual) statements, then it is a **contradiction** or **unsatisfiable**.

*Example*

- The composite statement $\lnot(P\rightarrow(Q\rightarrow(P\land Q)))$ is a contradiction.
- Verify this using a truth table with composite statement as  $\lnot(P\rightarrow(Q\rightarrow(P\land Q)))$

```{r boxC36, fig.cap='', out.width='100%', fig.asp=.7, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```


## Satisfiable

- If the truth values of a composite statement are sometimes true and sometimes false, irrespective of the truth values of the atomic statements, then it is called a satisfiable.


*Example*

- The composite statement $(P \rightarrow Q) \rightarrow (Q \rightarrow P)$ is satisfiable.
- Verify this using a truth table of $(P \rightarrow Q) \rightarrow (Q \rightarrow P)$

```{r boxC37, fig.cap='', out.width='100%', fig.asp=.7, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

## Duality Law

- Two formulae $P$ and $p^*$are said to be duals of each other if either one can be obtained from the other by interchanging $\land$ by $\lor$ and $\lor$ by $\land$.
- The two connectives $\lor$ and $\land$ are called dual to each other. 

*Example*

The formulae $P \equiv (P \lor Q)\land R$ and $P^* \equiv (P \land Q)\lor R$ are dual to each other.

\newpage

## Algebra of Propositions

- If $P, Q$ and $R$ be three statements, then the following laws hold. 

Algebra of Propositions|   
-----------------|-------------------
Commutative Laws  |  $P\land Q\equiv Q \land P$ and $P\lor Q\equiv Q \lor P$ 
Associative Laws  |  $P\land (Q \land R)\equiv (P\land Q) \land R$ and $P\lor (Q \lor R)\equiv (P\lor Q) \lor R$
Distributive Laws  |  $P\land (Q \lor R)\equiv (P\land Q) \lor (P\land R)$ and $P\lor (Q \land R)\equiv (P\lor Q) \land (P\lor R)$ 
Idempotent Laws  | $P\land P\equiv P$ and $P\lor P\equiv P$
Absorption Laws  |  $P\lor (P \land Q)\equiv P$ and $P\land (P \lor Q)\equiv P$ 
De Morgan's Laws  | $\lnot (P\land Q)  \Leftrightarrow (\lnot P) \lor(\lnot Q)$ and $\lnot (P\lor Q) \Leftrightarrow (\lnot P) \land (\lnot Q)$

## Logical Implication and Equivalence 
<!-- Schaum's outlines Boolean algebra pg 9-->

## Necessary and Sufficient Conditions
<!-- Refer diary  hand written 
note 2016  March14 -->

### Necessary Condition

### Sufficient Condition

### Necessary and Sufficient Condition

## References {-}

Acharjya, D. P. (2009). Fundamental approach to discrete mathematics. New Age International.

Mendelson, E. (1970). Boolean algebra and switching circuits. McGraw-Hill Edition 2004.

\newpage
\pagenumbering{arabic}
## Tutorial {-}

<!--schaum bool page 19 Question 1.1)-->
1. Reduce the following sentences to statement forms 

(a) A necessary condition for $x$ to be prime is that $x$ is odd or $x=2$.

(b) A sufficient condition for $f$ to be continuous is that $f$ is differentiable.

(c) A necessary and sufficient condition for Mr. Perera to be elected is that Mr. Perera wins 75 votes.

(d) Calathea plant will stay healthy only if enough moisture is available.

(e) It is raining but the sun is still shining.

(f) She will die today unless medical aid is obtained.

(g) If taxes are increased or government spending decreases, then inflation will not occur this year.

<!--Archariya page 7 1.11)-->
2. Show that the composite statement $(P \land (P \rightarrow Q)) \rightarrow Q$ is a tautology using a truth table


<!--Archariya page 7 1.12)-->
3. Show that the composite statement $\lnot (P \rightarrow( Q \rightarrow (P \land  Q)))$ is a contradiction using a truth table


<!--schaum bool page 9 example 1.19)-->
4. Show that $(A \rightarrow B) \rightarrow A$
logically implies $A$

<!--schaum bool page 9 example 1.20)-->
5. Show that $(A \leftrightarrow B)$ is 
logically equivalent to  $(A \rightarrow B) \land (B \rightarrow A)$

<!-- Fundamentals approach to Discrete mathematics book page 9-->
6. Find the negation of $A\rightarrow B$


<!-- Fundamentals approach to Discrete mathematics book page 9-->
7. Construct the truth table for $A \rightarrow ( B\leftrightarrow A \land B).$

<!-- Fundamentals approach to Discrete mathematics book page 10-->
8. Find the negation of the following statement "He is rich but unhappy"

<!-- Fundamentals approach to Discrete mathematics book page 10-->
9. Prove by constructing truth table
\[P \rightarrow(Q \lor R) \equiv (P \rightarrow Q) \lor (P \rightarrow R)\]

<!-- Fundamentals approach to Discrete mathematics book page 10-->
10. Find the negation of $P \leftrightarrow Q.$

11. With the help of the truth table prove that $\lnot (P \land Q) \equiv \lnot P\lor \lnot Q$

12. Show that $(P \rightarrow Q) \leftrightarrow (\lnot P \lor Q)$ is a tautology

13. Show that the following stamens are equivalent.

   - *Statement 1:* Good food is not cheap
   
   - *Statement 2:* Cheap food is not good

14. Express $P\rightarrow Q$ only using $\downarrow$ and $\uparrow$.

15. Prove that $(P \land Q) \land \lnot (P \lor Q)$ is a contradiction.

16. Express $P \leftrightarrow Q$ only using $\downarrow$ and $\uparrow$.

17. Show by truth table the following statement are equivalent

   - *Statement 1:* Rich men are unhappy
   
   - *Statement 2:* Men are unhappy or poor

<!-- Fundamentals approach to Discrete mathematics book page 13 Q 16-->
18. A constructor promises a client "We will fix it on Monday if it is not raining". When the constructor would be deemed to have broken his promise. Explain with the help of truth table. 

<!--schaum bool page 19 Question 1.2)-->
19. Eliminate as many parentheses as possible from

  a) $\{[(A \lor B)\rightarrow (\lnot C)] \lor [(((\lnot B)\land C)\land B)]\}$
  b) $\{[A \land (\lnot (\lnot B))]\leftrightarrow [B \leftrightarrow (C \lor B)]\}$
  c) $[(B\leftrightarrow (C \lor B)) \leftrightarrow (A \land (\lnot (\lnot B)))]$

<!--schaum bool page 20 Question 1.5)-->
20 Show the following are tautologies without using truth tables.

   a) $((A \rightarrow B) \rightarrow C) \rightarrow ((C \rightarrow A) \rightarrow(D \rightarrow A )).$
   b) $(A \rightarrow B)\rightarrow ((B \rightarrow C) \rightarrow (A \rightarrow C)).$