# Boolean Algebra

  - George Boole approached logic in a new way reducing it to a simple algebra.
  - He introduced symbolic logic known as *Boolean Algebra, Boolean function, Boolean expression, Boolean ring* etc.
  - Each variable in Boolean Algebra has either of two values: true or false.
  - The purpose of Boolean Algebra is to solve logic problems.
  - C.E Shannon observed that Boolean Algebra could be used to analyze electronic circuits.
  
## Gates

 - In Chapter 3 we discussed about logic connectives $\lnot$, $\land$ and $\lor$.
 - The connectives $\land$ and $\lor$ can be considered as circuits connected in series and parallel, respectively.
 -  A circuit with one or more input signals but only one output signal is known as a gate. 
 - Gates are digital circuits because of input and output signals, which are either low or high.
 - Gates are also known as logical circuits as they can be analyzed with Boolean Algebra.
 - In gates, the connectives $\lnot$, $\land$ and $\lor$ are usually denoted by the symbols $^\prime, \;\;.$ and $+$, respectively.
 
 <!-- the block diagrams fore the gates are discussed below-->
 
### NOT gate

A NOT gate receives input $x$, where $x$ is a bit (binary digit) and produces output $x^\prime$ where

\begin{equation}
x^\prime =
\begin{cases} 
1 & \text{if } x=0\\
0 & \text{if } x=1
\end{cases}
\end{equation}

  - $x^\prime$ is called the *complement* of $x$.
  - 0 is called the *zero* element.
  - 1 is called the *unit* element.
  - The output state is always the opposite of the input state.
  - The output is also known as the complement of the input.
  - The block diagram and the logic table for the basic NOT gate:

```{r boxC41, fig.cap='', out.width='100%', fig.asp=.7, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```
 
### AND gate

  - An AND gate receives input $x_1$ and $x_2$, where $x_1$ and $x_2$ are bits, and produces output $(x_1 \land x_2)$ where

\begin{equation}
(x_1 \land x_2) =
\begin{cases} 
1 & \text{if } x_1=x_2=1\\
0 & \text{ Otherwise }
\end{cases}
\end{equation}
  
  - $(x_1 \land x_2)$ is called the *meet* of $x_1$ and $x_2$.
  - An AND gate can have more than two inputs, but only one output.
 - The block diagram and the logic table for the basic AND gate:

```{r boxC42, fig.cap='', out.width='100%', fig.asp=.7, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

### OR gate

  - An OR gate receives input $x_1$ and $x_2$, where $x_1$ and $x_2$ are bits, and produces output $(x_1 \lor x_2)$ where

\begin{equation}
(x_1 \lor x_2) =
\begin{cases} 
1 & \text{if } x_1 \text{ or } x_2=1\\
0 & \text{ Otherwise }
\end{cases}
\end{equation}
  
  - $(x_1 \lor x_2)$ is called the *join* of $x_1$ and $x_2$.
  - An OR gate can have more than two inputs, but only one output.
  - The block diagram and the logic table for the basic OR gate:

```{r boxC43, fig.cap='', out.width='100%', fig.asp=.7, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

### More logic gates

- There are some other types of gates that are widely used in Computer Science such as NAND, NOR, XOR, and XNOR gates

#### NOR gate

 - A NOR gate receives input $x_1$ and $x_2$, where $x_1$ and $x_2$ are bits, and produces output $(x_1 \lor x_2)^\prime$ where

\begin{equation}
(x_1 \lor x_2)^\prime =
\begin{cases} 
1 & \text{if } x_1 = x_2=0\\
0 & \text{ Otherwise }
\end{cases}
\end{equation}
  
  - A NOR gate can have more than two inputs, but only one output.
  - The block diagram and the logic table for the basic NOR gate:

```{r boxC44, fig.cap='', out.width='100%', fig.asp=.7, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```


#### NAND gate

 - A NAND gate receives input $x_1$ and $x_2$, where $x_1$ and $x_2$ are bits, and produces output $(x_1 \land x_2)^\prime$ where

\begin{equation}
(x_1 \land x_2)^\prime =
\begin{cases} 
1 & \text{if } x_1 =0 \text{ or } x_2=0\\
0 & \text{ Otherwise }
\end{cases}
\end{equation}
  
  - A NAND gate can have more than two inputs, but only one output.
  - The block diagram and the logic table for the basic NAND gate:

```{r boxC45, fig.cap='', out.width='100%', fig.asp=.7, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

#### XOR gate (Exclusive OR gate)

 - A XOR  gate receives input $x_1$ and $x_2$, where $x_1$ and $x_2$ are bits, and produces output $(x_1 \overline{\lor} x_2)$ or $(x_1 \oplus x_2)$, where

\begin{equation}
(x_1 \oplus x_2)^\prime =
\begin{cases} 
1 & \text{if } x_1 =1 \text{ or } x_2=1 \text{ but not both}\\
0 & \text{ Otherwise }
\end{cases}
\end{equation}
  
  - Rule: XOR gate produces 1 that have an odd number of 1's.
  - A XOR gate can have more than two inputs, but only one output.
  - The block diagram and the logic table for the basic XOR gate:

```{r boxC46, fig.cap='', out.width='100%', fig.asp=.7, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```



#### XNOR gate (Exclusive NOR gate)

 - A XNOR  gate receives input $x_1$ and $x_2$, where $x_1$ and $x_2$ are bits, and produces output $(x_1\text{ XNOR }x_2)$, where

\begin{equation}
(x_1 \text{ XNOR } x_2)^\prime =
\begin{cases} 
1 & \text{if } x_1 \text{ and  } x_2 \text{ are same bits}\\
0 & \text{ Otherwise }
\end{cases}
\end{equation}

  - A XNOR gate can have more than two inputs, but only one output.
  - It can recognize *even-parity* words *i.e* a word which has an even number of 1's.
  - Example: 11001111 is even-parity as it contains six 1's, 1110 is an odd-parity as it has an odd number of 1's.
  - The block diagram and the logic table for the basic XNOR gate:

```{r boxC47, fig.cap='', out.width='100%', fig.asp=.7, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

## Combinatorial Circuit


  - A combinatorial circuit produces a unique output for every combination of inputs.
  - A combinatorial circuit has no memory, previous inputs and the state of the system do not affect the output of a combinatorial circuit.
  - These circuits can be constructed using gates which we have already discussed. 
  
```{r boxC48, fig.cap='', out.width='100%', fig.asp=.9, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```


\newpage

## Boolean Expression

  - Any expression built up from the variables $x_1,y_1,z_1, x_2,y_2,z_2,\dots$ by applying the operations $\land, \; \lor$ and $^\prime$ a finite number of times is called a Boolean expression.
  - If $X$ and $Y$ are two Boolean expressions, then $X^\prime, \;Y^\prime, \; (X\land Y)$ and $(X\lor Y)$ are also Boolean expressions.
  - The output of a combinatorial circuit is also a Boolean expression.
  
*Example*
  
```{r boxC49, fig.cap='', out.width='100%', fig.asp=.9, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

### Theorem
  
  - If $\land, \; \lor$ and $^\prime$ are connectives introduced earlier, then the following properties hold.
  
(i) Associative Law: For all $a,b,c \in \{0,1\}$
 \[(a \land b) \land c = a \land (b\land c) \text{ and } (a \lor b) \lor c = a \lor (b\lor c)\]
 
(ii) Identity Law: For all $a \in \{0,1\}$
 \[(a \land 1)  = a  \text{ and } (a \lor 0)  = a\]
 
(iii) Commutative Law: For all $a,b\in \{0,1\}$
 \[(a \land b)  = (b \land a)  \text{ and } (a \lor b)  = (b \lor a)\]
 
(iv) Complement Law: For all $a\in \{0,1\}$
 \[(a \land a^\prime)  = 0 \text{ and } (a \lor a^\prime)  = 1\]
 
(v) Distributive Law: For all $a,b,c\in \{0,1\}$
 \[a \lor (b\land c) = (a \lor b) \land (a \lor c)  \text{ and }  a \land (b\lor c)= (a \land b) \lor (a \land c)\]
 
(vi) De-Morgan's Law: If $x_1$ and $x_2$ are bits, *i.e.* $x_1, x_2\in \{0,1\},$ then
\[(x_1 \land x_2)^\prime = x_1^\prime \lor x_2^\prime  \text{ and }  (x_1 \lor x_2)^\prime = x_1^\prime \land x_2^\prime \] 

  
```{r boxC410, fig.cap='', out.width='100%', fig.asp=.9, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

## Equivalent Combinatorial Circuits

- Two combinatorial circuits, each having inputs $x_1, x_2, \dots, x_n$ are said to be equivalent if they produce the same output for same inputs.

```{r boxC411, fig.cap='', out.width='100%', fig.asp=.9, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))
```

\newpage

## Boolean Algebra

 - A *Boolean algebra*  consists of a set $S$ together with two binary operations $\land$ and $\lor$ on $S$, a singular operation $^\prime$ on $S$ and two specific elements 0 and 1 of $S$ such that the following laws hold.
 
  - A Boolean algebra will be designated by a hextuple $B = \langle S, \land, \lor, ^\prime, 0,1\rangle$
  - Sometimes one refers to the set $S$ as a Boolean algebra, but this is just a loose misuse of language.
  
(a) Associative Law: For all $a,b,c \in S$
 \[(a \land b) \land c = a \land (b\land c) \text{ and } (a \lor b) \lor c = a \lor (b\lor c)\]
 
(b) Identity Law: For all $a \in  S$
 \[(a \land 1)  = a  \text{ and } (a \lor 0)  = a\]
 
(c) Commutative Law: For all $a,b\in S$
 \[(a \land b)  = (b \land a)  \text{ and } (a \lor b)  = (b \lor a)\]
 
(d) Complement Law: For all $a\in S$
 \[(a \land a^\prime)  = 0 \text{ and } (a \lor a^\prime)  = 1\]
 
(e) Distributive Law: For all $a,b,c\in S$
 \[a \lor (b\land c) = (a \lor b) \land (a \lor c)  \text{ and }  a \land (b\lor c)= (a \land b) \lor (a \land c)\]
 
 
### Theorem

In a Boolean algebra: if $(a\lor b) =1$ and $(a\land b) =0,$ then $b=a^\prime,$ *i.e.* the complement is unique

```{r boxC412, fig.cap='', out.width='100%', fig.asp=.9, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))
```


### Theorem

In a Boolean algebra $B = \langle S, \land, \lor, ^\prime, 0,1\rangle$; the following properties hold.

  (a) Idempotent Law: For all $x \in S$
\[(x \lor x) =x \text{ and } (x \land x) =x\]
  (b) Bound Law: For all $x \in S$
\[(x \lor 1) =1 \text{ and } (x \land 0) =0\]
  (c) Absorption Law: For all $x,y \in S$
\[x \land (x\lor y) =x \text{ and } x \lor (x \land y) =x \]
  (d) Involution Law: For all $x \in S$
\[(x^\prime)^\prime  =x \]
  (e) 0 and 1 Law:  $0^\prime =1$ and $1^\prime =0$ 
  (f) De-Morgan's Law: For all $x, y \in S$
\[(x \land y)^\prime = x^\prime \lor y^\prime  \text{ and }  (x \lor y)^\prime = x^\prime \land y^\prime \] 

## Dual of a Statement

  -  The dual of a statement involving Boolean expressions is obtained by replacing 0 by 1, 1 by 0, $\land$ by $\lor$, and $\lor$ by $\land$.
  - Two Boolean expressions are said to be dual of each other if one expression is obtained from other by replacing 0 by 1, 1 by 0, $\land$ by $\lor$, and $\lor$ by $\land$.
  - In Boolean Algebra, the dual of a theorem is also a theorem.
  
*Example*

What the dual of the statement: $(x \land y)^\prime = x^\prime \lor y^\prime$

```{r boxC413, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))
```


## Boolean Function

Let $B = \langle S, \land, \lor, ^\prime, 0,1\rangle$  be a Boolean algebra and let $X(x_1, x_2,x_3, \dots, x_n)$ be a Boolean expression in $n$ variables.

A function $f: B^n \rightarrow B$ is called a Boolean function if $f$ is of the form

$f(x_1, x_2,x_3, \dots, x_n) = X(x_1, x_2,x_3, \dots, x_n)$

*Example*

Consider the Boolean function $f: B^3 \rightarrow B; \;\; B = \{0,1\}$ defined by 
\[f(x_!,x_2,x_3) = x_1 \land (x_2 \lor \bar{x_3})\]

```{r boxC414, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))
```

### Representation of Boolean Functions

- There are several ways of representing Boolean functions.
   
   (a) **Tabular representation**
   (b) $n$ Space representation
   (c) Cube representation
   
**Tabular representation**

  - A Boolean function is completely determined by its evaluation over any Boolean algebra.
  - In tabular representation, we consider a row $R$ of the table where the output is 1.
  - We then form the combination $(x_1 \land x_2 \land x_3 \land \dots \land x_n)$ and place a bar over each $x_i$ whose value is 0 in row $R$.
  - The combination formed is 1 if and only if $x_i$ have the value given in row $R$. 
  - We thus join (OR) the terms to obtain the Boolean expression.
  
\newpage 

*Example*

$x_1$ | $x_2$  | $x_3$  | $f(x_1, x_2, x_3)$
------|--------|--------|-------------------
  1   |   1    |   1    |    1
  1   |   1    |   0    |    0  
  1   |   0    |   1    |    1
  0   |   1    |   1    |    0  
  1   |   0    |   0    |    0
  0   |   1    |   0    |    1  
  0   |   0    |   1    |    0
  0   |   0    |   0    |    0  
  
```{r boxC415, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))
```  
  

## Various Normal Forms

### Disjunctive normal form

  - A Boolean function $f: B^n \rightarrow B$ which consists of a sum of elementary products is called the **disjunctive normal form** of the given function $f$
  - Let $f: B^n \rightarrow B$ is a Boolean function.
  - If $f$ is not identically zero, let $A_1, A_2, A_3, \dots, A_k$ denote the elements $A_i$ of $B_2^n,$ for which $f(A_i) =1,$ where

\[A_i = (a_1, a_2,\dots, a_n).\]

 - For each $A_i$ set $m_i = (y_1 \land y_2 \land y_3\land \dots \land y_n)$, where 
 
\begin{equation}
y_i =
\begin{cases} 
x_i & \text{if } a_i=1\\
x_i^\prime & \text{if } a_i=0.
\end{cases}
\end{equation}

 - Then $f(x_1, x_2, x_3, \dots, x_n)= m_1 \lor m_2 \lor m_3 \lor \dots \lor m_k.$
 - This representation of a Boolean function is called the **disjunctive normal form**
 
*Example*

Consider the Boolean function $(x_1 \oplus x_2).$ The truth table for this function is given below. 

$x_1$   |  $x_2$   |  $(x_1 \oplus x_2)$
--------|----------|----------------------
  1     |    1     |     0
  1     |    0     |     1
  0     |    1     |     1
  0     |    0     |     0
  
Write the disjunctive form of this function

```{r boxC416, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))
```
  
### Conjunctive normal form

  - A Boolean function $f: B^n \rightarrow B$ which consists of a product of elementary sums is called the **conjunctive normal form** of the given function $f$
  - Let $f: B^n \rightarrow B$ is a Boolean function.
  - If $f$ is not identically one, let $A_1, A_2, A_3, \dots, A_k$ denote the elements $A_i$ of $B_2^n,$ for which $f(A_i) =0,$ where

\[A_i = (a_1, a_2,\dots, a_n).\]

 - For each $A_i$ set $M_i = (y_1 \lor y_2 \lor y_3\lor \dots \lor y_n)$, where 
 
\begin{equation}
y_i =
\begin{cases} 
x_i & \text{if } a_i=0\\
x_i^\prime & \text{if } a_i=1.
\end{cases}
\end{equation}

 - Then $f(x_1, x_2, x_3, \dots, x_n)= M_1 \land M_2 \land M_3 \land \dots \land M_k.$
 - This representation of a Boolean function is called the **conjunctive normal form**
 
*Example*

Consider the Boolean function $(x_1 \oplus x_2).$ The truth table for this function is given below. 

$x_1$   |  $x_2$   |  $(x_1 \oplus x_2)$
--------|----------|----------------------
  1     |    1     |     0
  1     |    0     |     1
  0     |    1     |     1
  0     |    0     |     0
  
Write the conjunctive form of this function

```{r boxC417, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}
library(ggplot2)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "white", fill=NA, size=1))

```

- A term of the form $(y_1\land y_2 \land y3 \land \dots \land y_n),$  where each $y_i$ is either $x_i$ or $\bar{x_i}$ is called a *minterm*.
- A term of the form $(y_1 \lor y_2 \lor y3 \lor \dots \lor y_n),$  where each $y_i$ is either $x_i$ or $\bar{x_i}$ is called a *maxterm*

## References {-}

Acharjya, D. P. (2009). Fundamental approach to discrete mathematics. New Age International.

Mendelson, E. (1970). Boolean algebra and switching circuits. McGraw-Hill Edition 2004.