
# Descriptive Statistics {#intro}

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.path = 'figure/',
  cache.path = 'cache/',
  fig.align = 'center',
  fig.show = 'hold',
  cache = TRUE,
  external = TRUE,
  dev = "png",
  fig.height = 16,
  fig.width = 10,
  out.width = "0.8\\textwidth",
  fig.pos = "h"
)
```

\pagenumbering{arabic}

## 	Introduction to Statistics

###	Some Basic Terminologies Used in Statistics

**i	Population**

-	The set of **all** possible elements in the universe of interest to the researcher

**ii	Sample**

-	A Sample is a **subset** (a portion or part) of the population of interest
-	The sample must be a representative of the population of interest

**iii	Element**

-	Element is an **entity or object** which the information is collected.
-	*Eg: Student, household, farm, company, tomato plant*

**iv	Variable**

-	A variable is **a feature characteristic which has different ‘values’ or categories for different elements** (items/subjects/individuals)
-	*Eg: Gender of client, brand of mobile phones, risk level, number of emails received per day, age of client, income of client*

**v	Data**

-	Data are **measurements or facts** that are collected from a statistical unit/entity of interest
-	We collect data on variables
-	Data are raw numbers or facts that must be processed (analysed) to get useful information.
-	We get information from data.

-	*Eg:* 

  ***Variable:*** *Age (in years) of client*

  ***Data:*** *21, 45, 18, 32, 30, 22, 23, 27* 
    
  ***Information:***  

  *The mean age is 27.25 years*

  *The minimum age is 18 years*

  *The range of ages is 18-45*

  *The percentage of clients below 25 years of age: 50%*

**vi	Statistic**

-	**Characteristic** of a **sample**
-	The value which calculated based on sample data

**vii	Parameter**

-	**Characteristic** of a **population**
-	The value which calculated based on population data

**viii Census**

-	When a researcher **gathers data from the whole population for a given measurement,** it is called a census

**ix	Sampling**

-	When a researcher **gathers data from a sample of the population for a given measurement,** it is called sampling
-	The process of selecting a sample is also called sampling 


**Why take a sample instead of studying every member of the population ?**

-	Prohibitive cost of census
-	Destruction of item being studied may be required
-	Not possible to test or inspect all members of a population being studied.


###	Branches of Statistics 


```{r box1, fig.cap='', out.width='100%', fig.asp=.75, fig.align='center'}
#df<-data.frame(x = NULL, y = NULL)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

```

**i	Descriptive Statistics**

-	Descriptive statistics consists of organizing, summarizing and presenting data in an informative way.
-	The main purpose of descriptive statistics is to provide an overview of the data collected. 
-	Descriptive statistics describes the data collected through frequency tables, graphs and summary measures (mean, variance, quartiles, etc.).

**ii	Inferential Statistics**

-	In inferential statistics sample data are used to draw inferences (i.e. derive conclusions) or make predictions about the populations from which the sample has been taken.
-	This includes methods used to make decisions, estimates, predictions or generalizations about a population based on a sample. 
-	This includes point estimations, interval estimation, test of hypotheses, regression analysis, time series analysis, multivariate analysis, etc. 


### Types of Variables

```{r box2, fig.cap='', out.width='100%', fig.asp=.75, fig.align='center'}
#df<-data.frame(x = NULL, y = NULL)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

```


####	Qualitative / Quantitative Variables

**i Qualitative variable (Categorical variable)**

-	The characteristic is a quality.
-	The data are categories.
-	They cannot be given numerical values. 
-	However, it may be given a numerical label
-	Qualitative variables are sometimes referred as categorical variables.
-	*Eg:*

  *Gender:* 
  
  *Age group:*
  
  *Education level:* 
  
  *A/L stream:* 
  
  *Degree type:*

  *Hair colour: *  
   
  *FIT student batch:* 
  
  *Undergraduate level:*
  
  *Grade that you can obtain for CM 1110* 

**ii Quantitative variable **

-	The characteristic is a quantity
-	The data are numbers
-	Quantitative data require numeric values that indicate how much or how many.
-	They are obtained by counting or measuring with some scale
-	*Eg: *

  *Number of family members:* 
  
  *Number of emails received per day:* 
  
  *Weight of a student:*
  
  *Age:* 
  
  *Credit balance in the SIM card:*
  
  *Time remaining in class:*
  
  *Temperature:*
  
  *Marks *

####	Discrete/ Continuous Variables

-	Quantitative variables can be classified as either discrete or continuous.

**i Discrete Variables**

-	Quantitative
-	Usually the data are obtained by counting
-	There are impossible values between any two possible values 
-	*Eg:* 

   *Number of family members:*

   *Number of emails received per day:*

**ii Continuous Variables**

-	Quantitative
-	Usually, the data are obtained by measuring with a scale
-	There are no impossible values between any two possible values.(any value between any two possible values is also a possible value)
-	i.e a continuous variable can take any value within a specified range.
-	*Eg:*

   *Weight of a student:* 

   *Age:* 

   *Credit balance in the SIM card:* 
 
   *Time remaining in class:*

   *Temperature:*

   *Marks* 

###	Scales of Measurements


```{r box3, fig.cap='', out.width='100%', fig.asp=.75, fig.align='center'}
#df<-data.frame(x = NULL, y = NULL)

ggplot()+
  theme_void()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

```

-	There are four levels of measurements called, **nominal, ordinal, interval and ratio.**
-	Each levels has its own rules and restrictions
-	Different levels of measurement contains different amount of information with respect to whatever the data are measuring




**i	Nominal Scale**

-	Qualitative
-	No order or ranking in categories.
-	These categories have to be mutually exclusive, i.e. it should not be possible to place an individual or object in more than one category
-	A name of a category can be substituted by a number, but it will be mere label and have no numerical meaning

```{r box4, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}


ggplot()+
  theme_void()

```


**ii	Ordinal Scale**

-	Qualitative
-	Categories can be ordered or ranked
-	A name of a category can be substituted by a number, but such a sequence does not indicate absolute quantities.
-	Difference between any two numbers on the scale does not have a numerical meaningful.
-	It cannot be assumed that the differences between adjacent numbers on the scale are equal.

```{r box5, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}


ggplot()+
  theme_void()

```

**iii	Interval Scale**

-	Quantitative
-	Data can be ordered or ranked
-	There is no absolute zero point. Zero is only an arbitrary point with which other values can compare
-	Difference between two numbers is a meaningful numerical value
-	Ration of two numbers is not a meaningful numerical value.

```{r box6, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}


ggplot()+
  theme_void()

```

**iv	Ratio Scale**

-	Quantitative
-	Highest level of measurement
-	There exist an absolute zero point  (It has a true zero point)
-	Ratio between different measurements is meaningful

```{r box7, fig.cap='', out.width='100%', fig.asp=.5, fig.align='center'}


ggplot()+
  theme_void()

```

## Presentation of Data

### Tabular Presentations of Data 

### Graphic Presentations of Data
  
## Summary Measures

### Measures of Central Tendency
### Measures of Dispersion
### Measures of Skewness and Kurtosis
### Measures of Association